---
title: "DMR_finding_dmr"
author: "Chengzhou"
date: "2022-"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cluster the CpG

## DMR cluster



```{r warning=FALSE,message=FALSE}
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(minfi)
library(geograbi) 
library(limma)
library(corrplot)
library(maxLik)
```

## Function 

Define a function:

input: x, y are both n by p matrices 

output: a vector of length p indicating the column-wise correlation between x and y.


```{r}
colCors = function(x, y) {
        sqr = function(x) x*x
        if(!is.matrix(x)||!is.matrix(y)||any(dim(x)!=dim(y)))
                stop("Please supply two matrices of equal size.")
        x   = sweep(x, 2, colMeans(x))
        y   = sweep(y, 2, colMeans(y))
        cor = colSums(x*y) /  sqrt(colSums(sqr(x))*colSums(sqr(y)))
        return(cor)
}
```
## Function 
input:
Z: DNAm levels: n by L matrix; n is sample size; L is block size
X: covariates: n by p matrix; n is sample size; p is number of covariates; the first column must be phenotype
output
an maxLik object consisting of maximum likelihood, estimate, etc...
```{r}
get.mle.reg<-function(Z,X){
  X=as.matrix(X)
  Z=as.matrix(Z)
  p=ncol(X)-1
  n=nrow(Z)
  L=ncol(Z)
  log.Z=log(Z)
  log.1mZ=log(1-Z)
  log.denom=log(1+rowSums(Z/(1-Z)))
  starter.beta=mean((colMeans(Z)*(1-colMeans(Z))/apply(Z,MARGIN=2,var)-1)*(1-colMeans(Z)))
  if (p>0){
    starter.alpha=as.matrix(lm((log.Z-log.1mZ) ~ X[,2:(p+1)])[[1]])
    temp=rep(NA,length(p))
    for (i in 2:(p+1)){
      temp[i-1]=mean(starter.alpha[i,])
    }
    starter.alpha=c(starter.alpha[1,1:L],temp)
    starter=c(starter.alpha,starter.beta)
  }else{ 
    starter=c(as.vector(lm((log.Z-log.1mZ) ~ 1)[[1]]),starter.beta)
  }
  
  llk <- function(theta){
    coef.intercept=theta[1:L]
    if (p>0){
      coef.covariate=theta[(L+1):(L+p)]
      coef.beta=rbind(coef.intercept,matrix(coef.covariate,nrow=p,ncol=L))
      beta=theta[p+L+1]
      alpha=exp(X%*%coef.beta)*beta
    }else{
      coef.beta=matrix(coef.intercept,nrow=1)
      beta=theta[p+L+1]
      alpha=matrix(exp(coef.beta)*beta,nrow=n,ncol=L,byrow=TRUE)
    }
    if (L!=1){
      nom1=sum(lgamma(rowSums(alpha)+ beta))
      nom2=sum((alpha-1)*log.Z)-sum((alpha+1)*log.1mZ)
      denom1=sum(lgamma(alpha))+n*lgamma(beta)
      denom2=sum((rowSums(alpha)+beta)*log.denom)
      return (nom1+nom2-denom1-denom2)
    }else{
      return (sum(log(dbeta(Z,alpha,rep(beta,length(alpha))))))
    }
  }
  A=matrix(rep(0,p+L+1),nrow=1)
  A[p+L+1]=1
  B=as.matrix(0)
  res <- maxLik(llk, start = starter,constraints=list(ineqA=A, ineqB=B),iterlim=10000)
  return (res)
  
}
```
## Function 
output:joint p value of profile likelihood ratio/Wald method for that block (consiting of L CpG sites)
```{r}
ht<-function(Z,X.no.int){ #no intercept
  L=ncol(as.matrix(Z))
  X=cbind(1,X.no.int)
  res.constraint=get.mle.reg(Z,X[,-2]) #
  res.unconstraint=get.mle.reg(Z,X)
  p.lr=(1-pchisq(2*(res.unconstraint$maximum-res.constraint$maximum),1)) #
  p=ncol(as.matrix(X))
  index=L+1
  Fisher=-(res.unconstraint$hessian[index,index])
  p.Wald=1-pchisq(res.unconstraint$estimate[index]^2*Fisher,1)
  return (list(p.Wald=p.Wald, p.lr=p.lr))
  
}
```



```{r}
load('beta.Rdata')
t.beta <- t(beta) 
```
## preprocess the variables 

short the long row name of t.beta to identify with samples and vars

```{r}
rownames(t.beta) <- substring(rownames(t.beta),1,10)
colnames(beta) <- substring(colnames(beta),1,10)


#get the map information for CpG sites 
map.data.subset <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
colnames(map.data.subset)
#delete X and Y since our data delete the chr X and Chr Y 
map.data.subset = map.data.subset[map.data.subset$chr!='chrY'&map.data.subset$chr!='chrX',] 
map.data.subset <-map.data.subset$chr[map.data.subset$Name %in% rownames(beta)]
table(map.data.subset)

Z1 <- t.beta
dim(Z1)
```
## get the cluster information 

```{r}
start.loc.B=c()
end.loc.B=c()
count=0
for (chr.num in names(table((map.data.subset)))){
        count=count+1
        tbl=table((map.data.subset))
        
        which(names(tbl)==chr.num)
        index=(cumsum(tbl)[count]-tbl[count]+1):cumsum(tbl)[count]
        
        
        record_cor=colCors(Z1[,index[-length(index)]],Z1[,index[-1]])
        
        
        consecutive.table=rle(as.integer(record_cor>0.5))
        block.len=consecutive.table[[1]]
        block.val=consecutive.table[[2]]
        
        temp.loc=1:length(index)
        temp.block.start=rep(0,sum(block.val==1,na.rm=T))
        temp.block.end=rep(0,sum(block.val==1,na.rm=T))
        for (i in 1:sum(block.val==1,na.rm = T)){
                temp.block.start[i]=(cumsum(block.len)[block.val==1]-block.len[block.val==1]+1)[i]
                temp.block.end[i]=cumsum(block.len)[block.val==1][i]+1
                
        }
        
        temp.start.loc.B=1:length(index)
        temp.end.loc.B=1:length(index)
        if (sum(block.val==1,na.rm = T)>0){
                for (i in 1:sum(block.val==1,na.rm = T)){
                        temp.start.loc.B[temp.block.start[i]:temp.block.end[i]]=temp.block.start[i]
                        temp.end.loc.B[temp.block.start[i]:temp.block.end[i]]=temp.block.end[i]
                }
                temp.start.loc.B=unique(temp.start.loc.B)
                temp.end.loc.B=unique(temp.end.loc.B)
        }
        
        
        if (count==1){
                start.loc.B=temp.start.loc.B
                end.loc.B=temp.end.loc.B
        }else{
                start.loc.B=c(start.loc.B,temp.start.loc.B+cumsum(tbl)[count-1])
                end.loc.B=c(end.loc.B,temp.end.loc.B+cumsum(tbl)[count-1])
        }
        
}

table(end.loc.B-start.loc.B+1)
total_compare_LR <- sum(table(end.loc.B-start.loc.B+1))

L.vec <- as.numeric(names(table(end.loc.B-start.loc.B+1)))
length(L.vec)
```

## Extract the sammple information 

```{r}
# make all variables as numeric 
#the sample information 
samples <- geograbi.get.samples("GSE59065")
vars <- geograbi.extract.characteristics(samples)
rownames(vars) <- rownames(samples)
vars$`age (yr)` <- as.numeric(vars$`age (yr)`)
#rename the age column since this dataset the age is binary distribution
names(vars)[names(vars) == "age (yr)"] <- "age"

# make the age as categorical data
for (i in 1:nrow(vars)) {
        if(vars$age[i] > 50){
                vars$age[i]=1
        } else {
                vars$age[i]=0
        }
}
vars$age <- as.numeric(vars$age)
for (i in 1:nrow(vars)) {
        if(vars$gender[i] == "male"){
                vars$gender[i]=1
        } else {
                vars$gender[i]=0
        }
}
vars$gender <- as.numeric(vars$gender)

n=nrow(t.beta)

```


## estimate the parameters and conduct LR test to find the  DMR
```{r}
number.positive.cpg.LR <- rep(NA,length(L.vec))
count.L = 0
index1 <- NA
positi.record.LR1 <- NA
start.loc.B1 <- NA
end.loc.B1 <- NA
for (L in L.vec) {
        count.L = count.L +1
        
        positi.record.LR=rep(NA,length(which(end.loc.B-start.loc.B+1==L)))
        count=0
        for (i in which(end.loc.B-start.loc.B+1==L)){
                count=count+1
                index=start.loc.B[i]:end.loc.B[i]
                
                X = as.data.frame(cbind(vars$age,vars$gender))
                X= as.matrix(X,n,2) # two variables
                
                res=ht(t.beta[,index],X)
                positi.record.LR[count]=res$p.lr
                index2 <- index  
                index1 <- c(index1, index2)
                positi.record.LR1[index[1]] <- positi.record.LR[count]
                print(count)
        }
        number.positive.cpg.LR[count.L] <- L * length(positi.record.LR[positi.record.LR < 0.05/total_compare_LR])
}

# combine the position information and p-value
positi.record.LR2 <- positi.record.LR1[!is.na(positi.record.LR1)]
length(positi.record.LR2)
result_data_LR <- cbind(start.loc.B,end.loc.B,positi.record.LR2)
result_data_LR <- as.data.frame(result_data_LR)
colnames(result_data_LR) <- c("start","end","p_value")

save(result_data_LR,file = "result_data_LR.Rdata")

total_compare_LR
```

## extract the CpG information 
```{r}
DMR_LR <- result_data_LR[result_data_LR$p_value < 0.05/total_compare_LR,]
DMR_LR 
DMR_LR$n <- DMR_LR$end - DMR_LR$start + 1
total_cpg <- sum(DMR_LR$n)
total_cpg
posit <- NA
for (i in 1:nrow(DMR_LR)) {
        posit1 <- DMR_LR$start[i]:DMR_LR$end[i]
        posit <- c(posit,posit1)
        
}
posit <- posit[-1]
sig_cpg <- beta[posit,]
cpg_name_LR <- c(rownames(sig_cpg))
cpg_name_LR

```





